---
title: "Final Project"
author: "Volha Tryputsen"
date: "12/11/2018"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '6'
  pdf_document:
    highlight: tango
    toc: yes
    toc_depth: 6
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.height = 6, fig.width = 7)
```


```{r libraries}
library(tidyverse)
library(janitor)
library(knitr)
library(gee)
library(MuMIn)
library(forcats)
library(knitr)
library(lme4)
```


```{r data}
dat = 
  read.csv(file = '../frmgham2.csv') %>%
  clean_names()
```


```{r eval = FALSE}
dim(dat)

glimpse(dat)
head(dat)
View(dat)

dat %>% select(sex) %>% head()
```


```{r preprocessing}
apply(dat,2,function(x)100*sum(is.na(x))/11627)

dat = 
  dat %>%
  mutate(sex = factor(sex)) #%>% fct_recode(sex, "female" =  "2", "male" = "1")

# what variables have more than 70$ data missing
apply(dat,2,function(x)100*sum(is.na(x))/11627)[apply(dat,2,function(x)100*sum(is.na(x))/11627) > 70]

# exclude those variables( hdlc, ldlc)
dat = dat[,-c(22, 23)]


#center age for interpretation
hist(dat$age)
mean(dat$age)


dat = dat %>% mutate(cage = age - median(dat$age[dat$period == "1"]))
```




# EDA  
## response vs continuous covariates  
### (1): Smoking ~ age, sex   

```{r}

dat %>%
  select(cursmoke, age) %>% 
  ggplot(aes(x = age, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

dat %>%
  select(cursmoke, age) %>% 
  ggplot(aes(x = age, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  stat_summary(fun.y = 'mean', geom="line", col = 'red') +
  theme_bw() 
```

The proportion of smokers decreases with the age.


Smoking ~ age, sex  

```{r}
dat %>%
  select(cursmoke, age, sex) %>% 
  ggplot(aes(x = age, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```
There is a higher proportion of smoker among men compared to women as both age ,but there is no interaction between age and sex.  


### (2): Smoking ~ totchol, sex  

```{r}
dat %>%
  select(cursmoke, totchol) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = totchol, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```


smoking ~ totchol x sex [interaction]  
  
```{r}
dat %>%
  select(cursmoke, totchol, sex) %>% 
  ggplot(aes(x = totchol, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

Proportion of smokers has non-linear relationship with total cholesterol for women; proprtions increases with increase in total cholesterol for men (sex by totchol interaction effect). 


### (3): Smoking ~ sysbp, sex  
```{r}
dat %>%
  select(cursmoke, sysbp, sex) %>% 
  ggplot(aes(x = sysbp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


### (4): smoking ~ diabp, sex  
   
```{r}
dat %>%
  select(cursmoke, diabp, sex) %>% 
  ggplot(aes(x = diabp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```
Proportion of smokers decreases with increase of diastolic blood presure; the proprtions are higher for men (sex effect).  

### (5): smoking ~ cigpday, sex  
   
```{r}
dat %>%
  select(cursmoke, cigpday, sex) %>% 
  ggplot(aes(x = cigpday, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  #stat_summary(fun.y = 'mean', geom="line", col = 'red') +
  theme_bw() 
```


### (6): smoking ~ bmi, sex  
   
```{r}
dat %>%
  select(cursmoke, bmi, sex) %>% 
  ggplot(aes(x = bmi, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

dat %>%
  select(cursmoke, bmi) %>% 
  ggplot(aes(x = bmi, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  stat_summary(fun.y = 'mean', geom="line") +
  theme_bw() 
```

### (7): smoking ~ heartrte, sex  
   
```{r}
dat %>%
  select(cursmoke, heartrte, sex) %>% 
  ggplot(aes(x = heartrte, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (8): smoking ~ glucose, sex  
   
```{r}
dat %>%
  select(cursmoke, glucose, sex) %>% 
  ggplot(aes(x = glucose, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


### (9): smoking ~ timeap, sex  
   
```{r}
dat %>%
  select(cursmoke, timeap,sex) %>% 
  ggplot(aes(x = timeap, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (10): smoking ~ timemi, sex [not sig]
   
```{r}
dat %>%
  select(cursmoke, timemi,sex) %>% 
  ggplot(aes(x = timemi, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (11): smoking ~ timemifc, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timemifc,sex) %>% 
  ggplot(aes(x = timemifc, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (12): smoking ~ timemchd, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timechd,sex) %>% 
  ggplot(aes(x = timechd, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (13): smoking ~ timestrk, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timestrk,sex) %>% 
  ggplot(aes(x = timestrk, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (14): smoking ~ timecvd, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timecvd,sex) %>% 
  ggplot(aes(x = timecvd, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (14): smoking ~ timecvd, sex  [possibly?]
   
```{r}
dat %>%
  select(cursmoke, timecvd,sex) %>% 
  ggplot(aes(x = timecvd, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (15): smoking ~ timedth, sex  [possibly?]
   
```{r}
dat %>%
  select(cursmoke, timedth,sex) %>% 
  ggplot(aes(x = timedth, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (16): smoking ~ timehyp, sex  [possibly?]
   
```{r}
dat %>%
  select(cursmoke, timehyp,sex) %>% 
  ggplot(aes(x = timehyp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```




## response vs categorical covariates  

### smoking ~ sex  
   
```{r}
c(chisq.test(table(dat$sex[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$sex, dat$cursmoke))$p.value)
```

### smoking ~ diabetes  
   
```{r}
c(chisq.test(table(dat$diabetes[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$diabetes, dat$cursmoke))$p.value)
```

### smoking ~ bpmeds  
   
```{r}
c(chisq.test(table(dat$bpmeds[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$bpmeds, dat$cursmoke))$p.value)
```


### smoking ~ educ 
   
```{r}
c(chisq.test(table(dat$educ[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$educ, dat$cursmoke))$p.value)
```

### smoking ~ prevchd 
   
```{r}
c(chisq.test(table(dat$prevchd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevchd, dat$cursmoke))$p.value)
```

### smoking ~ prevap 
   
```{r}
c(chisq.test(table(dat$prevap[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevap, dat$cursmoke))$p.value)
```


### smoking ~ prevmi [not sig]  
   
```{r}
c(chisq.test(table(dat$prevmi[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevmi, dat$cursmoke))$p.value)
```

### smoking ~ prevstrk [sig all]  
   
```{r}
c(chisq.test(table(dat$prevstrk[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevstrk, dat$cursmoke))$p.value)
```


### smoking ~ prevhyp [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$prevhyp[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevhyp, dat$cursmoke))$p.value)
```

### smoking ~ death [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$death[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$death, dat$cursmoke))$p.value)
```

### smoking ~ death [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$death[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$death, dat$cursmoke))$p.value)
```

### smoking ~ angina [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$angina[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$angina, dat$cursmoke))$p.value)
```


### smoking ~ mi_fchd  
   
```{r}
c(chisq.test(table(dat$mi_fchd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$mi_fchd, dat$cursmoke))$p.value)
```

### smoking ~ anychd [not sig]  
   
```{r}
c(chisq.test(table(dat$anychd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$anychd, dat$cursmoke))$p.value)
```

### smoking ~ stroke [not sig]  
   
```{r}
c(chisq.test(table(dat$stroke[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$stroke, dat$cursmoke))$p.value)
```

### smoking ~ cvd [borderline sig]  
   
```{r}
c(chisq.test(table(dat$cvd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$cvd, dat$cursmoke))$p.value)
```

### smoking ~ hyperten [highly sig]  
   
```{r}
c(chisq.test(table(dat$hyperten[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$hyperten, dat$cursmoke))$p.value)
```




## pairwise correlations  
```{r}
dat[,-c(1,2, 20, 38)] %>%
  cor(., use = "complete.obs") %>%
  corrplot::corrplot(., type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 75, tl.offset = 1, tl.cex = .8, method = "ellipse")
```





## demographics  
### education  
```{r}
dat %>%
  group_by(educ) %>%
  summarise(mean_resp = mean(cursmoke),
            lower = prop.test(sum(cursmoke),n())$conf.int[1],
            upper = prop.test(sum(cursmoke),n())$conf.int[2]) %>%
  ggplot(aes(x = educ, y = mean_resp)) +
  geom_errorbar(aes(ymin=lower, ymax=upper), width=.1) +
  geom_line() +
  theme_bw() +
  labs(title = 'Proportion of smokers by education (mean and 95% CI)', caption = "Figure 2")
```



# Model  
## Smoking ~ age+ sex + covariates  


```{r}

dat %>% 
  na.omit() %>%
  group_by(randid) %>% 
  mutate(num_visits = n()) %>% 
  ungroup() %>% 
  dplyr::select(randid, num_visits) %>% 
  distinct() %>% 
  summarise(mean(num_visits), sd(num_visits), 
            max(num_visits), min(num_visits), median(num_visits)) %>%
  kable(caption = "Number of observations per subject stats") %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = F, position = "left")


dat %>% 
  filter(time>0) %>%
  ggplot(aes(x = as.factor(period), y = time)) + 
  geom_boxplot()

```

Motivation for conditional model: 
- Incomplete data - imbalanced number of observations per person: after missing data is removed the mean/median number of observations per subject is 2.2/2 with sd = 0.7. The minimum number of visits is 1 and the maximum is 3 (**Table above**).  
- Unequally spaced f/up time: the spread of a distribution of number of days from the 1 st visit is within 200 days (**Boxplot above**).  


### demographics  
```{r}
m1 = lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + (1|randid), family = binomial, data = dat, nAGQ = 0)
m2 = lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + (1|randid), family = binomial, data = dat, nAGQ = 0)

summary(m1)
summary(m2)

AIC(m1, m2)
# m2 is better 
```

### heart  
```{r}
###
# adding heart variables: heartrte, prevstrk, totchol
m3a = lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + heartrte + (1|randid), family = binomial, data = dat, nAGQ = 0)

# m3b didnt converge
m3b= lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + as.factor(prevstrk)  + (1|randid), family = binomial, data = dat, nAGQ = 0)


m3c =
  dat %>%
  select(randid, cursmoke, age, sex, educ, bmi, totchol) %>%
  na.omit() %>%
  lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + log(totchol) + (1|randid), family = binomial, data = ., nAGQ = 0)


AIC(m2, m3a, m3b, m3c)

summary(m3a)
summary(m3b)
summary(m3c)

# totchol has the best AIC 
```


### blood presure - nothing is significant   
Check blood presure predictors: systolic, diastolic bp, bp medications, hypertantion, prevhyp    

```{r}
# dichotomize education into high school and college 
dat$educ_binary <- NA
dat$educ_binary[dat$educ %in% c(1,2)] <- "high_school"
dat$educ_binary[dat$educ %in% c(3,4)] <- "college"

table(dat$educ)
table(dat$educ_binary)

dat = dat %>% mutate(educ_binary = factor(educ_binary))

m3c =
  dat %>%
  select(randid, cursmoke, age, sex, educ, bmi, totchol) %>%
  na.omit() %>%
  lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + log(totchol) + (1|randid), family = binomial, data = ., nAGQ = 0)

m3c1 =
  dat %>%
  select(randid, cursmoke, age, sex, educ_binary, bmi, totchol) %>%
  na.omit() %>%
  lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + (1|randid), family = binomial, data = ., nAGQ = 0)


summary(m3c1)
AIC(m3c, m3c1)


m4a= lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + sysbp + (1|randid), family = binomial, data = dat, nAGQ = 0)


m4b= lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + diabp + (1|randid), family = binomial, data = dat, nAGQ = 0)


m4c= lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + hyperten + (1|randid), family = binomial, data = dat, nAGQ = 0)

m4d= lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + prevhyp + (1|randid), family = binomial, data = dat, nAGQ = 0)

m4e= lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + bpmeds + (1|randid), family = binomial, data = dat, nAGQ = 0)

AIC(m4a, m4b, m4c, m4d, m4e)
summary(m4a)
summary(m4b)
summary(m4c)
summary(m4d)
summary(m4e)

plot(log(dat$totchol), dat$diabp)

#hist(dat$sysbp)
hist(log(dat$sysbp))
hist(dat$diabp)
#hist(log(dat$diabp))
```


### diabetes  - nothing  
Check diabetes and glucose  
```{r}
#start model
m3c1 =
  dat %>%
  select(randid, cursmoke, age, sex, educ_binary, bmi, totchol) %>%
  na.omit() %>%
  lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + (1|randid), family = binomial, data = ., nAGQ = 0)

m5a = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + diabetes + (1|randid), family = binomial, data = dat, nAGQ = 0)

m5b = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + log(glucose) + (1|randid), family = binomial, data = dat, nAGQ = 0)


AIC(m5a, m5b)
summary(m5a)
summary(m5b)

hist(dat$glucose)
hist(log(dat$glucose))
```


## Prevalence   
```{r}
m3c1 =
  dat %>%
  select(randid, cursmoke, age, sex, educ_binary, bmi, totchol) %>%
  na.omit() %>%
  lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + (1|randid), family = binomial, data = ., nAGQ = 0)

m6a = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + mi_fchd + (1|randid), family = binomial, data = dat, nAGQ = 0)


m6b = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + prevchd +  (1|randid), family = binomial, data = dat, nAGQ = 0)

m6c = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + prevap + (1|randid), family = binomial, data = dat, nAGQ = 0)

m6d = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + death + (1|randid), family = binomial, data = dat, nAGQ = 0)

m6e = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + angina + (1|randid), family = binomial, data = dat, nAGQ = 0)

m6f = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + hospmi + (1|randid), family = binomial, data = dat, nAGQ = 0)

m6g = lme4::glmer(cursmoke ~ age + sex + educ_binary + bmi + log(totchol) + cvd + (1|randid), family = binomial, data = dat, nAGQ = 0)

AIC(m6a, m6b, m6c, m6d, m6e, m6f, m6g)
summary(m6a) # mi_fchd
summary(m6b) # prevchd
summary(m6c) # prevap
summary(m6d) # death
summary(m6f) # hospmi
summary(m6g) # cvd

summary(m6e) # angina(no)



m6h = lme4::glmer(cursmoke ~ cage + sex + educ_binary + bmi + log(totchol) + prevchd + death + cvd + (1|randid), family = binomial, data = dat, nAGQ = 0)

summary(m6h)
```


Final model:  
```{r}
m6h = lme4::glmer(cursmoke ~ cage + sex + educ_binary + bmi + log(totchol) + prevchd + death + cvd + (1|randid), family = binomial, data = dat, nAGQ = 0)

summary(m6h)
```



## Number of cigarets ~ age + sex + covariates  


## systolic bp ~ smoking+ covariates  


## diastolic bp ~ smoking + covariates  


## total cholesterol ~ smoking + covariates  



