---
title: "Final Project"
author: "Volha Tryputsen"
date: "12/01/2018"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '6'
  pdf_document:
    highlight: tango
    toc: yes
    toc_depth: 6
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.height = 6, fig.width = 7)
```


```{r libraries}
library(tidyverse)
library(janitor)
library(knitr)
library(gee)
library(MuMIn)
library(forcats)
library(knitr)



library(RLRsim) # for testing significance of random intercept
```


```{r data}
dat = 
  read.csv(file = '../frmgham2.csv') %>%
  clean_names()
```


```{r eval = FALSE}
dim(dat)

glimpse(dat)
head(dat)
View(dat)

dat %>% select(sex) %>% head()
```


```{r preprocessing}
dat = 
  dat %>%
  mutate(sex = factor(sex)) #%>% fct_recode(sex, "female" =  "2", "male" = "1")
```


# Smoking vs. Age, Sex
## (1): Smoking ~ age, sex  
### Is there a relationship between age and smoking status? 
ANS: Yes, the proportion of smokers decreases with the age.  

```{r}
dat %>%
  select(cursmoke, age) %>% 
  ggplot(aes(x = age, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


### Does this relationship differ by sex?  
ANS: There is a higher proportion of smoker among men compared to women as both age ,but there is no interaction between age and sex.      
```{r}
dat %>%
  select(cursmoke, age, sex) %>% 
  ggplot(aes(x = age, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


## (2) number of sigarets  ~ age, sex 
### Is there a relationship between the number of cigarettes smoked per day and age?   
ANS: Yes, number of sigarets smoked per day stays constant for 30-50 years old and decreases with age after 50 years old.   

#### All 

```{r}
dat %>%
  select(cigpday, age) %>% 
  ggplot(aes(x = age, y = cigpday)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


#### Smokers only  

```{r}
dat %>%
  select(cigpday, age) %>% 
  filter(cigpday > 0) %>%
  ggplot(aes(x = age, y = cigpday)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```



### Does this relationship differ by sex?  
ANS: There is sex effect (men smoke higer number of sigarets per day than women across age), but there is no sex and age interaction.  

#### All  
```{r}
dat %>%
  select(cigpday, age, sex) %>% 
  ggplot(aes(x = age, y = cigpday, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

#### Smokers only    
```{r}
dat %>%
  select(cigpday, age, sex) %>% 
  filter(cigpday > 0) %>% 
  ggplot(aes(x = age, y = cigpday, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```



# Smoking vs. health outcomes  

## (1) The relationship between current smoking status and systolic blood pressure.  
### smoking ~ sysbp  
ANS: Proportion of smokers decreases with increase of systolic blood presure  
```{r}
dat %>%
  select(cursmoke, sysbp) %>% 
  ggplot(aes(x = sysbp, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: slightly higher sysbp for non-smokers  
```{r}
dat %>%
  select(cursmoke, sysbp) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = sysbp, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```

### smoking ~ sysbp, sex  
ANS: Proportion of smokers decreases with increase of systolic blood presure; the proportion is higher for men (sex effect).     
```{r}
dat %>%
  select(cursmoke, sysbp, sex) %>% 
  ggplot(aes(x = sysbp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no differences in sysbp between male and female smokers and non-smokers  
```{r}
dat %>%
  select(cursmoke, sex, sysbp) %>% 
  mutate(cursmoke = factor(cursmoke),
           smoke_sex = interaction(cursmoke, sex)) %>% 
  ggplot(aes(y = sysbp, x = smoke_sex)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```


## (2) The relationship between current smoking status and diastolic blood pressure.  

### smoking ~ diabp  
ANS: Proportion of smokers decreases with increase of diastolic blood presure for BP=100 ad then proportion increases again (latter could be due to not enough data)  
```{r}
dat %>%
  select(cursmoke, diabp) %>% 
  ggplot(aes(x = diabp, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference    
```{r}
dat %>%
  select(cursmoke, diabp) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = diabp, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```

### smoking ~ diabp, sex  
ANS: Proportion of smokers decreases with increase of diastolic blood presure; the proprtions are higher for men (sex effect).     
```{r}
dat %>%
  select(cursmoke, diabp, sex) %>% 
  ggplot(aes(x = diabp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference  
```{r}
dat %>%
  select(cursmoke, sex, diabp) %>% 
  mutate(cursmoke = factor(cursmoke),
           smoke_sex = interaction(cursmoke, sex)) %>% 
  ggplot(aes(y = diabp, x = smoke_sex)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```


## (3) The relationship between current smoking status and serum total cholesterol.  
### smoking ~ totchol   
ANS: Proportion of smokers slightly decreases with increase of total cholesterol values   
```{r}
dat %>%
  select(cursmoke, totchol) %>% 
  ggplot(aes(x = totchol, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference    
```{r}
dat %>%
  select(cursmoke, totchol) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = totchol, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```


### smoking ~ totchol, sex [!!!]  
ANS: Proportion of smokers hasnonlinera relationship with total cholesterol for women; proprtions increases with increase in total cholesterol for men (sex by totchol interaction effect).     
```{r}
dat %>%
  select(cursmoke, totchol, sex) %>% 
  ggplot(aes(x = totchol, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference  
```{r}
dat %>%
  select(cursmoke, sex, totchol) %>% 
  mutate(cursmoke = factor(cursmoke),
           smoke_sex = interaction(cursmoke, sex)) %>% 
  ggplot(aes(y = totchol, x = smoke_sex)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```



