---
title: "Final Project"
author: "Volha Tryputsen"
date: "12/11/2018"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '6'
  pdf_document:
    highlight: tango
    toc: yes
    toc_depth: 6
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.height = 6, fig.width = 7)
```


```{r libraries}
library(tidyverse)
library(janitor)
library(knitr)
library(gee)
library(MuMIn)
library(forcats)
library(knitr)
library(lme4)
```


```{r data}
dat = 
  read.csv(file = '../frmgham2.csv') %>%
  clean_names()
```


```{r eval = FALSE}
dim(dat)

glimpse(dat)
head(dat)
View(dat)

dat %>% select(sex) %>% head()
```


```{r preprocessing}
dat = 
  dat %>%
  mutate(sex = factor(sex)) #%>% fct_recode(sex, "female" =  "2", "male" = "1")

# what variables have more than 70$ data missing
apply(dat,2,function(x)100*sum(is.na(x))/11627)[apply(dat,2,function(x)100*sum(is.na(x))/11627) > 70]

# exclude those variables( hdlc, ldlc)
dat = dat[,-c(22, 23)]
```




# EDA  
## response vs continuous covariates  
### (1): Smoking ~ age, sex   

```{r}
dat %>%
  select(cursmoke, age) %>% 
  ggplot(aes(x = age, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```
The proportion of smokers decreases with the age.


Smoking ~ age, sex  

```{r}
dat %>%
  select(cursmoke, age, sex) %>% 
  ggplot(aes(x = age, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```
There is a higher proportion of smoker among men compared to women as both age ,but there is no interaction between age and sex.  


### (2): Smoking ~ totchol, sex  

```{r}
dat %>%
  select(cursmoke, totchol) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = totchol, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  geom_jitter(width = 0.1, alpha = 0.1) +
  theme_bw() 
```


smoking ~ totchol x sex [interaction]  
  
```{r}
dat %>%
  select(cursmoke, totchol, sex) %>% 
  ggplot(aes(x = totchol, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

Proportion of smokers has non-linear relationship with total cholesterol for women; proprtions increases with increase in total cholesterol for men (sex by totchol interaction effect). 


### (3): Smoking ~ sysbp, sex  
```{r}
dat %>%
  select(cursmoke, sysbp, sex) %>% 
  ggplot(aes(x = sysbp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


### (4): smoking ~ diabp, sex  
   
```{r}
dat %>%
  select(cursmoke, diabp, sex) %>% 
  ggplot(aes(x = diabp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```
Proportion of smokers decreases with increase of diastolic blood presure; the proprtions are higher for men (sex effect).  

### (5): smoking ~ cigpday, sex  
   
```{r}
dat %>%
  select(cursmoke, cigpday, sex) %>% 
  ggplot(aes(x = cigpday, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


### (6): smoking ~ bmi, sex  
   
```{r}
dat %>%
  select(cursmoke, bmi, sex) %>% 
  ggplot(aes(x = bmi, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (7): smoking ~ heartrte, sex  
   
```{r}
dat %>%
  select(cursmoke, heartrte, sex) %>% 
  ggplot(aes(x = heartrte, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (8): smoking ~ glucose, sex  
   
```{r}
dat %>%
  select(cursmoke, glucose, sex) %>% 
  ggplot(aes(x = glucose, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


### (9): smoking ~ timeap, sex  
   
```{r}
dat %>%
  select(cursmoke, timeap,sex) %>% 
  ggplot(aes(x = timeap, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (10): smoking ~ timemi, sex [not sig]
   
```{r}
dat %>%
  select(cursmoke, timemi,sex) %>% 
  ggplot(aes(x = timemi, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (11): smoking ~ timemifc, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timemifc,sex) %>% 
  ggplot(aes(x = timemifc, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (12): smoking ~ timemchd, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timechd,sex) %>% 
  ggplot(aes(x = timechd, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 

```

### (13): smoking ~ timestrk, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timestrk,sex) %>% 
  ggplot(aes(x = timestrk, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (14): smoking ~ timecvd, sex  [not sig]
   
```{r}
dat %>%
  select(cursmoke, timecvd,sex) %>% 
  ggplot(aes(x = timecvd, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (14): smoking ~ timecvd, sex  [possibly?]
   
```{r}
dat %>%
  select(cursmoke, timecvd,sex) %>% 
  ggplot(aes(x = timecvd, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (15): smoking ~ timedth, sex  [possibly?]
   
```{r}
dat %>%
  select(cursmoke, timedth,sex) %>% 
  ggplot(aes(x = timedth, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### (16): smoking ~ timehyp, sex  [possibly?]
   
```{r}
dat %>%
  select(cursmoke, timehyp,sex) %>% 
  ggplot(aes(x = timehyp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```




## response vs categorical covariates  

### smoking ~ sex  
   
```{r}
c(chisq.test(table(dat$sex[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$sex, dat$cursmoke))$p.value)
```

### smoking ~ diabetes  
   
```{r}
c(chisq.test(table(dat$diabetes[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$diabetes, dat$cursmoke))$p.value)
```

### smoking ~ bpmeds  
   
```{r}
c(chisq.test(table(dat$bpmeds[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$bpmeds, dat$cursmoke))$p.value)
```


### smoking ~ educ 
   
```{r}
c(chisq.test(table(dat$educ[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$educ, dat$cursmoke))$p.value)
```

### smoking ~ prevchd 
   
```{r}
c(chisq.test(table(dat$prevchd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevchd, dat$cursmoke))$p.value)
```

### smoking ~ prevap 
   
```{r}
c(chisq.test(table(dat$prevap[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevap, dat$cursmoke))$p.value)
```


### smoking ~ prevmi [not sig]  
   
```{r}
c(chisq.test(table(dat$prevmi[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevmi, dat$cursmoke))$p.value)
```

### smoking ~ prevstrk [sig all]  
   
```{r}
c(chisq.test(table(dat$prevstrk[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevstrk, dat$cursmoke))$p.value)
```


### smoking ~ prevhyp [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$prevhyp[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$prevhyp, dat$cursmoke))$p.value)
```

### smoking ~ death [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$death[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$death, dat$cursmoke))$p.value)
```

### smoking ~ death [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$death[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$death, dat$cursmoke))$p.value)
```

### smoking ~ angina [1/2 sig]  
   
```{r}
c(chisq.test(table(dat$angina[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$angina, dat$cursmoke))$p.value)
```


### smoking ~ mi_fchd  
   
```{r}
c(chisq.test(table(dat$mi_fchd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$mi_fchd, dat$cursmoke))$p.value)
```

### smoking ~ anychd [not sig]  
   
```{r}
c(chisq.test(table(dat$anychd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$anychd, dat$cursmoke))$p.value)
```

### smoking ~ stroke [not sig]  
   
```{r}
c(chisq.test(table(dat$stroke[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$stroke, dat$cursmoke))$p.value)
```

### smoking ~ cvd [borderline sig]  
   
```{r}
c(chisq.test(table(dat$cvd[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$cvd, dat$cursmoke))$p.value)
```

### smoking ~ hyperten [highly sig]  
   
```{r}
c(chisq.test(table(dat$hyperten[dat$period == 1], dat$cursmoke[dat$period == 1]))$p.value, 
  chisq.test(table(dat$hyperten, dat$cursmoke))$p.value)
```




## pairwise correlations  
```{r}
dat[,-c(1,2, 20)] %>%
  cor(., use = "complete.obs") %>%
  corrplot::corrplot(., type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 75, tl.offset = 1, tl.cex = .8, method = "ellipse")
```



# Model  
```{r}
m1 = lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + (1|randid), family = binomial, data = dat)
m2 = lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + (1|randid), family = binomial, data = dat)

# m3a didnt converge
m3a = lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + heartrte + (1|randid), family = binomial, data = dat)
# m3b didnt converge
m3b= lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + as.factor(prevstrk)  + (1|randid), family = binomial, data = dat)


m3c =
  dat %>%
  select(cursmoke, age, sex, educ, bmi, totchol) %>%
  na.omit() %>%
  lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + log(totchol) + (1|randid), family = binomial, data = .)



m4 = lme4::glmer(cursmoke ~ age + sex + as.factor(educ) + bmi + (1|randid), family = binomial, data = dat)

AIC(m2, m3)

summary(m2)

apply(dat,2,function(x)100*sum(is.na(x))/11627)
```



