---
title: "Untitled"
output: pdf_document
---
```{r}
library(foreign)
library(tidyr)
library(dplyr)
library(ggplot2)
library(lme4)
library(nlme)
library(RLRsim)
library(CompRandFld)
library(MASS)
library(gee)
library(geepack)
library(VIM)
library(Amelia)
library(reshape2)

#import the data
my.data <- read.csv("F:/courses/longitudinal/FINAL/frmgham2.csv")
head(my.data)
summary(my.data$AGE)

as.data.frame(table(my.data$SEX))
```

(1) Is there a relationship between age and smoking status? 
Does this relationship differ by sex?
```{r}
model.q1 <- glmer(CURSMOKE ~ AGE + as.factor(SEX) + AGE:as.factor(SEX) + as.factor(educ) 
                 + BMI + DIABETES + HEARTRTE + PREVCHD + PREVSTRK 
                 + PREVHYP + TIMEDTH + (1|RANDID),
                 data = my.data,
                 family=binomial,
                 corstr = "exchangeable")
summary(model.q1)
exp(coef(summary(model.q1)))
```
We can see from the model that with age increases 1, the log odds ratio of smoking against non-smoking is -0.051.

Which means that with age incraeses 1, the odds ratio of smoking against non-smoking is 1.0523. The odds of currently smoking is 5% lower than the odds of currently not smoking with the age incraeses 1 unit.

We can see that both age and sex have a p-value less than 0.05. Thus, sex can differ the relationship between age and smoking status. For women, the log odds ratio of smoking against non-smoking is -0.76 as the age increases 1. Thus, for women, the odds of smoking against non-smoking is 53.2% lower with the age increases 1 unit.

(2) Is there a relationship between the number of cigarettes smoked per day and age? 
Does this relationship differ by sex?
```{r}
model.q2 <- lmer(CIGPDAY ~ AGE + as.factor(SEX) + as.factor(educ) + (1|RANDID),
                  data = my.data)
summary(model.q2)
```
If we think cig per day as count data, it follows poisson distribution.
Then we can fit GEE model as well:
```{r}
model.q2_1 <- glmer(CIGPDAY ~ AGE + as.factor(SEX) + AGE:as.factor(SEX) + as.factor(educ) +BMI + DIABETES + HEARTRTE + PREVCHD + PREVSTRK + (1|RANDID) 
                 + PREVHYP + TIMEDTH,
                  data = my.data,
                  na.action = "na.omit",
                  family=poisson,
                  corstr = "exchangeable")
summary(model.q2_1)
exp(coef(summary(model.q2_1)))
```
From the result we can see that the p-value of age and sex are both show significance. Thus, sex can differ the relationship between cigraettes per day.

From the model that with age increases 1, the number of cigarettes per day will decrease 2.32% at a population level for men.

For women, cigaretees per day will decrease 55.9% with one unit of age increases at a population level.

