---
title: "LDA final project"
author: "Kaitlin Maciejewski, Morgan de Ferrante, Bingnan Li, Volha Tryputsen"
output:
  # html_document:
  #   df_print: paged
  #   toc: yes
  #   toc_depth: '6'
  #   toc_float: true
  #   code_folding: hide
  pdf_document:
    highlight: tango
    toc: yes
    toc_depth: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, out.width = "70%", cache = T)
library(tidyverse)
library(janitor)
library(knitr)
library(MuMIn)
library(forcats)
library(RLRsim)
data <- read.csv("../final_data/frmgham2.csv") %>% clean_names()
attach(data)
```

# Exploratory

## Smoking vs. Age, Sex
## (1): Smoking ~ age, sex  
### Is there a relationship between age and smoking status? 
ANS: Yes, the proportion of smokers decreases with the age.  

```{r, smoke_v_age}
data %>%
  select(cursmoke, age) %>% 
  ggplot(aes(x = age, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


### Does this relationship differ by sex?  
ANS: There is a higher proportion of smoker among men compared to women as both age ,but there is no interaction between age and sex.      
```{r, smoke_age_sex}
data %>%
  select(cursmoke, age, sex) %>% 
  ggplot(aes(x = age, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


## (2) number of cigarettes  ~ age, sex 
### Is there a relationship between the number of cigarettes smoked per day and age?   
ANS: Yes, number of sigarets smoked per day stays constant for 30-50 years old and decreases with age after 50 years old.   

#### All 

```{r, n_c_age_all}
data %>%
  select(cigpday, age) %>% 
  ggplot(aes(x = age, y = cigpday)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```


#### Smokers only  

```{r, n_c_age_smoke}
data %>%
  select(cigpday, age) %>% 
  filter(cigpday > 0) %>%
  ggplot(aes(x = age, y = cigpday)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

### Does this relationship differ by sex?  
ANS: There is sex effect (men smoke higer number of sigarets per day than women across age), but there is no sex and age interaction.  

#### All  
```{r, n_c_age_s_all}
data %>%
  select(cigpday, age, sex) %>% 
  ggplot(aes(x = age, y = cigpday, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

#### Smokers only    
```{r, n_c_age_s_smoke}
data %>%
  select(cigpday, age, sex) %>% 
  filter(cigpday > 0) %>% 
  ggplot(aes(x = age, y = cigpday, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

## Smoking vs. health outcomes  

## (1) The relationship between current smoking status and systolic blood pressure.  
### smoking ~ sysbp  
ANS: Proportion of smokers decreases with increase of systolic blood presure  
```{r, smoke_sbp}
data %>%
  select(cursmoke, sysbp) %>% 
  ggplot(aes(x = sysbp, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: slightly higher sysbp for non-smokers  
```{r, smoke_sysbp_status}
data %>%
  select(cursmoke, sysbp) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = sysbp, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  theme_bw() 
```

### smoking ~ sysbp, sex  
ANS: Proportion of smokers decreases with increase of systolic blood presure; the proportion is higher for men (sex effect).     
```{r, smoke_sysbp_sex}
data %>%
  select(cursmoke, sysbp, sex) %>% 
  ggplot(aes(x = sysbp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no differences in sysbp between male and female smokers and non-smokers  
```{r, smoke_sysbp_sex_status}
data %>%
  select(cursmoke, sex, sysbp) %>% 
  mutate(cursmoke = factor(cursmoke),
           smoke_sex = interaction(cursmoke, sex)) %>% 
  ggplot(aes(y = sysbp, x = smoke_sex)) +
  geom_boxplot(outlier.colour = "white") +
  theme_bw() 
```


## (2) The relationship between current smoking status and diastolic blood pressure.  

### smoking ~ diabp  
ANS: Proportion of smokers decreases with increase of diastolic blood presure for BP=100 ad then proportion increases again (latter could be due to not enough data)  
```{r, smoke_dbp}
data %>%
  select(cursmoke, diabp) %>% 
  ggplot(aes(x = diabp, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference    
```{r, smoke_dbp_box}
data %>%
  select(cursmoke, diabp) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = diabp, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  theme_bw() 
```

### smoking ~ diabp, sex  
ANS: Proportion of smokers decreases with increase of diastolic blood presure; the proprtions are higher for men (sex effect).     
```{r, smoke_dbp_s}
data %>%
  select(cursmoke, diabp, sex) %>% 
  ggplot(aes(x = diabp, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference  
```{r, smoke_dbp_s_bp}
data %>%
  select(cursmoke, sex, diabp) %>% 
  mutate(cursmoke = factor(cursmoke),
           smoke_sex = interaction(cursmoke, sex)) %>% 
  ggplot(aes(y = diabp, x = smoke_sex)) +
  geom_boxplot(outlier.colour = "white") +
  theme_bw() 
```


## (3) The relationship between current smoking status and serum total cholesterol.  
### smoking ~ totchol   
ANS: Proportion of smokers slightly decreases with increase of total cholesterol values   
```{r, smoke_tc}
data %>%
  select(cursmoke, totchol) %>% 
  ggplot(aes(x = totchol, y = cursmoke)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference    
```{r, smoke_tc_bp}
data %>%
  select(cursmoke, totchol) %>% 
  mutate(cursmoke = factor(cursmoke)) %>%
  ggplot(aes(y = totchol, x = cursmoke)) +
  geom_boxplot(outlier.colour = "white") +
  theme_bw() 
```


### smoking ~ totchol, sex [!!!]  
ANS: Proportion of smokers hasnonlinera relationship with total cholesterol for women; proprtions increases with increase in total cholesterol for men (sex by totchol interaction effect).     
```{r, smoke_tc_sex}
data %>%
  select(cursmoke, totchol, sex) %>% 
  ggplot(aes(x = totchol, y = cursmoke, group = sex, color = sex)) +
  geom_jitter(height = 0.1, alpha = 0.1) +
  geom_smooth(lwd = 1.5) +
  theme_bw() 
```

ANS: no difference  
```{r, smoke_tc_sex_bp}
data %>%
  select(cursmoke, sex, totchol) %>% 
  mutate(cursmoke = factor(cursmoke),
           smoke_sex = interaction(cursmoke, sex)) %>% 
  ggplot(aes(y = totchol, x = smoke_sex)) +
  geom_boxplot(outlier.colour = "white") +
  theme_bw() 
```

## Cor plot
```{r}
library(dplyr)

corr <- data[,c(-1,-21)] %>% cor(., use = "complete.obs") 

library(corrplot)

corrplot(corr, type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 75, 
         tl.offset = 1, tl.cex= .8, method = "ellipse")
```


## Missingness
```{r, missingness}
prop <- round(colSums(is.na(data))/dim(data)[1], 3)

knitr::kable(sort(prop, decreasing = TRUE)[1:9], col.names = "Proportion of NAs")

#MISSING DATA ANALYSIS
VIM::aggr(data, prop=T, numbers=T)
mice::md.pattern(data)
# HDLC and LDLC have a lot of missing data, and these variables should be eliminated

prob.data <- data %>%
  group_by(period) %>% 
  summarise(sysbp_prob = sum(sysbp, na.rm = TRUE)/n())
prob.data

table(data$period)
```

```{r, eval = F, echo = F}
##plot these probabilities 
ggplot(prob.data, aes(y=sysbp_prob,
                      x=period)) + 
  geom_line() + 
  labs(color = 'Avg Systolic BP')


prob.data <- data %>%
  group_by(period) %>% 
  summarise(diabp_prob = sum(diabp, na.rm = TRUE)/n())

##plot these probabilities 
ggplot(prob.data, aes(y=diabp_prob,
                      x=period)) + 
  geom_line() + 
  labs(color = 'Avg Diastolic BP')

prob.data <- data %>%
  group_by(period) %>% 
  summarise(totchol_prob = sum(totchol, na.rm = TRUE)/n())

##plot these probabilities 
ggplot(prob.data, aes(y=totchol_prob,
                      x=period)) + 
  geom_line() + 
  labs(color = 'Avg Total Chol BP')


cdplot(factor(cursmoke) ~ sysbp, data=data, 
       ylab = "Current Smoking Status", 
       xlab = "Systolic BP")

cdplot(factor(cursmoke) ~ diabp, data=data, 
       ylab = "Current Smoking Status", 
       xlab = "Diatolic BP")

cdplot(factor(cursmoke) ~ totchol, data=data, 
       ylab = "Current Smoking Status", 
       xlab = "Total Cholesterol")
```

# Models

```{r}
library(lme4)
library(dplyr)
```

(1) Is there a relationship between age and smoking status? Does this relationship differ by sex?
```{r}
summary(gee::gee(cursmoke ~ age + sex + as.factor(educ) + bmi + diabetes+ heartrte + prevchd + prevstrk + prevhyp + timedth, id = randid, family=binomial, corstr = "unstructured"))
```

(2) Is there a relationship between the number of cigarettes smoked per day and age? Does this
relationship differ by sex?
```{r}
summary(gee::gee(cigpday ~ age + sex + as.factor(educ),  id = randid, corstr = "unstructured")) 
```

While answering these questions, please account for any confounders that you have evidence may impact the relationship between age and sex with smoking.

Next you are interested in the relationship between certain health outcomes and smoking status. In particular you are interested in:

(1) The relationship between current smoking status and systolic blood pressure. 
```{r}
summary(gee::gee(sysbp ~ cursmoke + age + sex + as.factor(educ), id = randid, na.action = "na.omit", corstr = "unstructured"))
```

(2) The relationship between current smoking status and diastolic blood pressure. 
```{r}
summary(gee::gee(diabp ~ cursmoke + age + sex + as.factor(educ), id = randid, na.action = "na.omit", corstr = "unstructured"))
```

(3) The relationship between current smoking status and serum total cholesterol.
```{r}
summary(gee::gee(totchol ~ cursmoke + age + sex + as.factor(educ), id = randid, na.action = "na.omit", corstr = "unstructured"))
```

```{r}
model_1<-glmer(cursmoke~ age + sex + sysbp + diabp + totchol + as.factor(educ) + (1|randid), family=binomial, na.action = "na.omit")

knitr::kable(summary(model_1)$coefficients,digits = 3)
```

```{r, eval = F}
glmer(cursmoke~ age + sex + totchol + as.factor(educ) + bmi + diabetes+ heartrte + prevchd + prevstrk + prevhyp + timedth + (1|randid), family=binomial, na.action = "na.omit") %>% knitr::kable(summary(.)$coefficients,digits = 3) # too many covariates to converge
```

# models from bingnan ...

## model of EDA:
(1) Is there a relationship between age and smoking status? 
Does this relationship differ by sex?
```{r}
my.data <- read.csv("../final_data/frmgham2.csv")
library(gee)

model.q1 <- gee(CURSMOKE ~ AGE + as.factor(SEX) + as.factor(educ) 
                 + BMI + DIABETES + HEARTRTE + PREVCHD + PREVSTRK 
                 + PREVHYP + TIMEDTH, 
                 id = RANDID,
                 data = my.data,
                 family=binomial, 
                 corstr = "unstructured")
summary(model.q1)
```
(2) Is there a relationship between the number of cigarettes smoked per day and age? 
Does this relationship differ by sex?
```{r}
model.q2 <- lmer(CIGPDAY ~ AGE + as.factor(SEX) + as.factor(educ) + (1|RANDID),
                  data = my.data)
summary(model.q2)
```
If we think cig per day as count data, it follows poisson distribution.
Then we can fit GEE model as well:
```{r}
model.q2_1 <- gee(CIGPDAY ~ AGE + as.factor(SEX) + as.factor(educ),
                  data = my.data,
                  id = RANDID,
                  family=poisson,
                  corstr = "unstructured")
summary(model.q2_1)
```

## model including age and sex
(1) The relationship between current smoking status and systolic blood pressure.
```{r}
model.p1 <- gee(SYSBP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID, 
                data = my.data,
                na.action = "na.omit",
                corstr = "unstructured")
summary(model.p1)
#fit exchangeable model
model.p1.1 <- gee(SYSBP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID, 
                data = my.data,
                na.action = "na.omit",
                corstr = "exchangeable")
summary(model.p1.1)
```
Compare the naive SE and robust SE we can see that exchangeable model is reasonable.


(2) The relationship between current smoking status and diastolic blood pressure. 
```{r}
model.p2 <- gee(DIABP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID,
                data = my.data,
                na.action = "na.omit", 
                corstr = "unstructured")
summary(model.p2)
#fit exchangeable model
model.p2.2 <- gee(DIABP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID,
                data = my.data,
                na.action = "na.omit", 
                corstr = "exchangeable")
summary(model.p2.2)
```
(2) The relationship between current smoking status and serum total cholesterol.
```{r}
model.p3 <- gee(TOTCHOL ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ),
                data = my.data,
                id = RANDID, 
                na.action = "na.omit", 
                corstr = "unstructured")
summary(model.p3)
#fit exchangeable model
model.p3.3 <- gee(TOTCHOL ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ),
                data = my.data,
                id = RANDID, 
                na.action = "na.omit", 
                corstr = "exchangeable")
summary(model.p3.3)
```

Using mixed effect model using cig per day instead of smoking status:
```{r}
#saturated model
model.saturated <- lmer(CIGPDAY ~ as.factor(SEX) + AGE 
                            + BPMEDS + as.factor(educ)
                            + TOTCHOL + BMI + GLUCOSE + DIABETES + HEARTRTE + PREVAP
                            + PREVCHD + PREVMI + PREVSTRK +STROKE+ PREVHYP + (1|RANDID),
                              na.action = 'na.omit',
                              data = my.data)
summary(model.saturated)
```

```{r}
#using variables that selected
model.mixed2 <- lmer(CIGPDAY~ AGE + as.factor(SEX) + SYSBP 
                      + DIABP + TOTCHOL + as.factor(educ) 
                      + (1|RANDID),
                      data = my.data,
                      na.action = "na.omit")
summary(model.mixed2)
```

# models from morgan

```{r}
library(dplyr)

# data %>%
#   group_by(randid) %>%
#   summarise(sum_death = sum(death)) %>%
#   filter(sum_death > 0 )

# library(lme4)
# 
# final_data <- data %>%
#   select(cursmoke, age, sex, educ,totchol,bmi,diabetes,heartrte,prevchd, prevstrk, prevhyp, timedth , randid) %>%
#   mutate(sex = factor(sex),
#          educ = factor(educ))
# 
# fit <- glmer(cursmoke ~ age + factor(sex) + factor(educ)  + bmi + 
#              diabetes + heartrte + prevchd + prevstrk + prevhyp  + (1 | randid),
#              family = binomial, 
#              na.action = "na.omit") 
# 
# 
# fit <- gee(cursmoke ~ age + factor(sex) + factor(educ)  + bmi + 
#                diabetes + heartrte + prevchd + prevstrk + prevhyp + timedth,
#            id = randid,
#            family = "binomial",
#            na.action = "na.omit")

##################
# Initial Models #
##################


# totchol_fit <- gee(totchol ~ cursmoke + age + factor(sex) + factor(educ)  + bmi + 
#              diabetes + heartrte + prevchd + prevhyp +  prevstrk + death,
#            id = randid,
#            family = "gaussian",
#            na.action = "na.omit")
# 
# round(2 * pnorm(abs(coef(summary(totchol_fit))[,5]), lower.tail = FALSE), 3) 
# 
# 
# sysbp_fit <- gee(sysbp ~ cursmoke + age + factor(sex) + factor(educ)  + bmi + 
#                      diabetes + heartrte + prevchd + prevstrk + death,
#                    id = randid,
#                    family = "gaussian",
#                    na.action = "na.omit")
# round(2 * pnorm(abs(coef(summary(sysbp_fit))[,5]), lower.tail = FALSE), 3) 
# 
# diabp_fit <- gee(diabp ~ cursmoke + age + factor(sex) + factor(educ)  + bmi + 
#                      diabetes + heartrte + prevchd + prevstrk +death,
#                    id = randid,
#                    family = "gaussian",
#                    na.action = "na.omit")
# 
# round(2 * pnorm(abs(coef(summary(diabp_fit))[,5]), lower.tail = FALSE), 3) 


###############################################
# Models After Removing Non Significant Terms #
###############################################

totchol_fit <- gee(totchol ~ cursmoke + age + factor(sex) + bmi + 
                     diabetes + heartrte  + prevhyp ,
                   id = randid,
                   family = "gaussian",
                   corstr = "unstructured",
                   na.action = "na.omit")

knitr::kable(summary(totchol_fit)$coefficients[,c(1,4,5)], digits = 3)
knitr::kable(round(2 * pnorm(abs(coef(summary(totchol_fit))[,5]), lower.tail = FALSE), 3)) 
QIC(totchol_fit)

sysbp_fit <- gee(sysbp ~ cursmoke + age + factor(sex)  + bmi + 
                   diabetes + heartrte + prevchd + prevstrk + death,
                 id = randid,
                 family = "gaussian",
                 corstr = "unstructured",
                 na.action = "na.omit")

knitr::kable(summary(sysbp_fit)$coefficients[,c(1,4,5)], digits = 3)
knitr::kable(round(2 * pnorm(abs(coef(summary(sysbp_fit))[,5]), lower.tail = FALSE), 3))
QIC(sysbp_fit)

diabp_fit <- gee(diabp ~ cursmoke + factor(sex) + factor(educ) + bmi +
                   diabetes + heartrte  + prevstrk +death,
                 id = randid,
                 family = "gaussian",
                 corstr = "unstructured",
                 na.action = "na.omit")

knitr::kable(summary(diabp_fit)$coefficients[,c(1,4,5)], digits = 3)
knitr::kable(round(2 * pnorm(abs(coef(summary(diabp_fit))[,5]), lower.tail = FALSE), 3))
QIC(diabp_fit)

```
