---
title: "Kaitlin's EDA"
author: "Kaitlin Maciejewski"
date: "12/1/2018"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# Models GLMER

(1) Is there a relationship between age and smoking status? Does this relationship differ by sex?

(2) Is there a relationship between the number of cigarettes smoked per day and age? Does this relationship differ by sex?

(1) The relationship between current smoking status and systolic blood pressure. 

(2) The relationship between current smoking status and diastolic blood pressure. 

(3) The relationship between current smoking status and serum total cholesterol.
```{r}
library(lme4)
library(dplyr)
data <- read.csv("../final_data/frmgham2.csv") %>% janitor::clean_names() 

attach(data)

model_1<-glmer(cursmoke~ age + sex + sysbp + diabp+ totchol + as.factor(educ) + (1|randid), family=binomial, na.action = "na.omit")

knitr::kable(summary(model_1)$coefficients,digits = 3)
```

```{r}
glmer(cursmoke~ age + sex + totchol + as.factor(educ) + bmi + diabetes+ heartrte + prevchd + prevstrk + prevhyp + timedth + (1|randid), family=binomial, na.action = "na.omit") %>% knitr::kable(summary(.)$coefficients,digits = 3)
```