---
title: "LDA final project"
output: pdf_document
---
```{r}
library(foreign)
library(tidyr)
library(dplyr)
library(ggplot2)
library(lme4)
library(nlme)
library(RLRsim)
library(CompRandFld)
library(MASS)
library(gee)
library(geepack)
library(VIM)
library(Amelia)
library(reshape2)

#import the data
my.data <- read.csv("../final_data/frmgham2.csv")
head(my.data)
summary(my.data$AGE)
```

EDA of relationship between age and smoking status
```{r}
boxplot.1 <- ggplot(data = my.data, aes(x = CURSMOKE, y = AGE, group=CURSMOKE))+
  geom_boxplot()+
  ggtitle('Boxplot of smoking status against age')
boxplot.1
boxplot.2 <- ggplot(data = my.data, aes(x = CURSMOKE, y = AGE, group=CURSMOKE))+
  geom_boxplot()+
  facet_wrap(~ SEX,ncol=2)+
  ggtitle('Boxplot of smoking status against age grouped by sex')
boxplot.2
```

EDA of relationship between number of cigarettes smoked per day and age
```{r}
spaghettiplot.1 <- ggplot(data=my.data, aes(x = AGE, y = CIGPDAY))+
  geom_point()+
  geom_smooth(method='loess')+
  ggtitle('association between age and cig per day')
spaghettiplot.1
spaghettiplot.2 <- ggplot(data=my.data, aes(x = AGE, y = CIGPDAY))+
  geom_point()+
  geom_smooth(method='loess')+
  facet_wrap(~SEX, ncol=2)+
  ggtitle('association between age and cig per dat grouped by sex')
spaghettiplot.2
```

BPMEDS adjusted
```{r}
BPMEDS <- ggplot(data=my.data, aes(x = AGE, y = CIGPDAY, group = RANDID))+
  geom_point()+
  facet_wrap(~BPMEDS, ncol=2)+
  ggtitle('association between age and cig per dat grouped by BPMEDS')
BPMEDS
```
We find that there are some missing data in BPMEDS.

MISSING DATA ANALYSIS
```{r}
VIM::aggr(my.data, prop=T, numbers=T)
mice::md.pattern(my.data)
```

#model of EDA:
(1) Is there a relationship between age and smoking status? 
Does this relationship differ by sex?
```{r}
model.q1 <- gee(CURSMOKE ~ AGE + as.factor(SEX) + as.factor(educ) 
                 + BMI + DIABETES + HEARTRTE + PREVCHD + PREVSTRK 
                 + PREVHYP + TIMEDTH, 
                 id = RANDID,
                 data = my.data,
                 family=binomial, 
                 corstr = "unstructured")
```
(2) Is there a relationship between the number of cigarettes smoked per day and age? 
Does this relationship differ by sex?
```{r}
model.q2 <- lmer(CIGPDAY ~ AGE + as.factor(SEX) + as.factor(educ) + (1|RANDID),
                  data = my.data)
summary(model.q2)
```
If we think cig per day as count data, it follows poisson distribution.
Then we can fit GEE model as well:
```{r}
model.q2_1 <- gee(CIGPDAY ~ AGE + as.factor(SEX) + as.factor(educ),
                  data = my.data,
                  id = RANDID,
                  family=poisson,
                  corstr = "unstructured")
```

#model including age and sex
(1) The relationship between current smoking status and systolic blood pressure.
```{r}
model.p1 <- gee(SYSBP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID, 
                data = my.data,
                na.action = "na.omit",
                corstr = "unstructured")
summary(model.p1)
#fit exchangeable model
model.p1.1 <- gee(SYSBP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID, 
                data = my.data,
                na.action = "na.omit",
                corstr = "exchangeable")
summary(model.p1.1)
```
Compare the naive SE and robust SE we can see that exchangeable model is reasonable.


(2) The relationship between current smoking status and diastolic blood pressure. 
```{r}
model.p2 <- gee(DIABP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID,
                data = my.data,
                na.action = "na.omit", 
                corstr = "unstructured")
summary(model.p2)
#fit exchangeable model
model.p2.2 <- gee(DIABP ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ), 
                id = RANDID,
                data = my.data,
                na.action = "na.omit", 
                corstr = "exchangeable")
summary(model.p2.2)
```
(2) The relationship between current smoking status and serum total cholesterol.
```{r}
model.p3 <- gee(TOTCHOL ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ),
                data = my.data,
                id = RANDID, 
                na.action = "na.omit", 
                corstr = "unstructured")
summary(model.p3)
#fit exchangeable model
model.p3.3 <- gee(TOTCHOL ~ CURSMOKE + AGE + as.factor(SEX) 
                + as.factor(educ),
                data = my.data,
                id = RANDID, 
                na.action = "na.omit", 
                corstr = "exchangeable")
summary(model.p3.3)
```

Using mixed effect model using cig per day instead of smoking status:
```{r}
#saturated model
model.saturated <- lmer(CIGPDAY ~ as.factor(SEX) + AGE 
                            + BPMEDS + as.factor(educ)
                            + TOTCHOL + BMI + GLUCOSE + DIABETES + HEARTRTE + PREVAP
                            + PREVCHD + PREVMI + PREVSTRK +STROKE+ PREVHYP + (1|RANDID),
                              na.action = 'na.omit',
                              data = my.data)
summary(model.saturated)
```

```{r}
#using variables that selected
model.mixed2 <- lmer(CIGPDAY~ AGE + as.factor(SEX) + SYSBP 
                      + DIABP + TOTCHOL + as.factor(educ) 
                      + (1|RANDID),
                      data = my.data,
                      na.action = "na.omit")
summary(model.mixed2)
```








